Rethinking how students learn to design programs

### 2016-04-23 manuscript-body.tex
adjusted first person; added note about first person in interview section

What's common to these threads of research is how they mine the data. In
most applications of code snapshot\footnote{What I call ``code
  snapshot'' research may be more formally called ``online protocol''
  research \cite{jadud_methods_2006,rodrigo_analyzing_2009}. ``Online'' here isn't mean to mean
  the sense of globally-connected computers, but rather that student
  materials are collected in a minimally-intrusive way \emph{while
  students work}.} research, the aim is to average across events and
sessions to arrive at a characteristic measure of a student's behavior.
Jadud \cite{jadud_methods_2006}, for example, develops the idea of a student's ``Error
Quotient,'' a measure of how effectively a student addresses
compile-time errors in their code.\footnote{A useful way of thinking
  about Jadud's \cite{jadud_methods_2006} error quotient is that the computation believes
  in second chances. It doesn't penalize students for making errors; it
  penalizes students if they don't fix a given error before compiling
  the same piece of code \emph{again.}} Rodrigo and colleagues extend
the idea of error quotient and introduce similar
measures such as students' ``compilation profiles'' \cite{rodrigo_analyzing_2009}, ``frustration profiles'' \cite{rodrigo_affective_2009}, and ``error
profiles'' \cite{tabanao_predicting_2011}. Because these measures can be
collected in real-time as students code, they offer the potential to
identify at-risk students and respond with early
interventions.

@@

The data presented in this case are taken from an ongoing IRB-approved
study of undergraduate electrical engineering majors undertaken at
\emph{Flagship State}, a large, public research institution on the east
coast. For two semesters, I have followed a total of 10 students taking
``Intermediate Programming Concepts for Engineers.'' It is the second of
a required two-semester course sequence in programming.\footnote{Hereafter,
  I use ``Intermediate Programming'' to refer to the second course and
  ``Basic Programming'' to refer to the first.} Students can (and some
do) place out of Basic Programming via AP Computer Science credit, but
all Electrical and Computer Engineering students must take Intermediate
Programming.

@@

My analysis details the work of Rebecca, a female first-year electrical
engineering major. I focus specifically on Rebecca's code for finding
``one-stop'' flights, which the instructor defined as ``all pairs of
flights that route the user between the departure and arrival airports
with exactly 1 stop (i.e., a one-connection flight)'' (Flights Database
handout, 2012). The one-stop problem is particularly challenging. To
solve it successfully students' code must accept a user's choice of
airports and day, then stitch together routes that involve two separate
flights in a way that passes stringent constraints for acceptable
layover times.

@@

This study proceeds from an empirical challenge: how can we develop
accounts of students' programming activity that explain the form and
evolution of their code on a design project? The focused form of that
challenge for this study is ``how can we understand the unconventional
design choices embedded in Rebecca's one-stop flight code?'' To answer
that question, my study draws from three data streams: ethnographic
observation, clinical interviewing, and code snapshot analysis.

For two semesters, I ethnographically embedded myself in the same
instructor's section of Intermediate Programming. My aim throughout was
to see what students see in terms of course material, assignment
directives, and instruction. In fall 2011 I observed approximately 50\%
of the course lectures. I also independently completed all class
homeworks and three out of the four course projects to more fully
understand the course's assessments. In spring 2012 I continued
attending lectures, though less frequently, and began attending select
TA-led discussion sections. During both lectures and discussion sections
I took field notes while recording ambient audio using a LiveScribe
Pulse pen.

Rebecca was one of four students (three female, one male) willing and
able to participate in a series of 1-hour outside-of-class clinical
interviews during the spring 2012 semester.\footnote{In total, roughly
  30 students responded to my initial in-class solicitation to be
  contacted by email about my study. Of the students I emailed,
  approximately 8 students responded to my emails to schedule interview
  times. Of those 8, only four students were able to successfully find
  interview time slots that fit our respective schedules.} I interviewed
Rebecca five times, and in typical interviews I split time between
asking about her experiences in the course and giving her time in the
interview to work on her project code. During each interview, I
simultaneously used:

@@

The final component of data gathering is modeled after Jadud's (2006)
system for capturing students' code. My colleagues and I developed
software, built around the open-source version control system called
Git, that effectively creates an entire copy of a student's code---what
we call ``snapshots''---every time students invoke the compiler on their
code. Our software then sends those snapshots to a secure,
researcher-accessible server in real-time as they're created.
Consequently, I could plan each interview with Rebecca around
up-to-the-minute knowledge of her work---in some cases work she had
completed just hours before the interview---and tailor my interview
questions to emerging patterns in her code. In total, Rebecca's work
resulted in 958 compilation snapshots over the course of the semester.

@@

In this phase of analysis, I restrict data to only the final version of \texttt{one\_stop\_flight.c} Rebecca submitted as part of her project. This restriction is important because of the data that gets left out. We're forced to see Rebecca's design work the way her instructor did when he graded it: as a final product. The submitted code contains little---if any---evidence of design iteration. And, we have almost no access to streams of activity (inscriptional, gestural, verbal, and otherwise material) that would help us understand Rebecca's early stage design \citep{petre_software_2014}. Indeed, other than the code itself, the only artifacts that actually carry Rebecca's voice in this phase of analysis are the comments she places in her code. Put another way: restricting our scope to just code---and only the final submitted code at that---denies the analyst access to the channels of ``talk, embodied action, and inscription'' involved in design work \citep[ p.~179]{hall_disrupting_2002}.

@@

While Rebecca's design isn't optimized for speed, it uses drastically less memory than do other solutions (including the instructor's official solution). What Rebecca could have done (and what I'll discuss in a later section) is read the entire contents of all data files into RAM. I'll call that approach an \emph{in-memory} solution, because all of the flight information is loaded into memory. Instead, her solution stores only a handful of lines (4 at most) in memory at a time, leaving the rest \emph{on-disk}. An analogy helps clarify the difference.

@@

For each of the four projects during the semester, the instructor gave students what he called a ``design brief.'' Each design brief outlined the problem to be solved as well as any constraints imposed on students' solutions. For example, in the flights project, students' programs had to reject candidate multi-stop trip routes if the layover time would be too short (under 30 minutes) or too long (more than two hours). But, in addition to what I might call \emph{user-centered constraints} (viz., reject multi-stop trips that would have grueling layovers), the instructor also directed students on \emph{implementation details}: ways their program should work at a technical level that would be invisible to the user.

@@

A second key feature of Rebecca's code is the almost identical repetition of a single 23-line code chunk seven times (lines 50--240). Because users can run queries by choosing a day to fly (and some flights only run on certain days), students' code must be able to handle each of the seven possible days for when a user would want to fly. In principle, Rebecca's code achieves just that.\footnote{I say ``in principle'' because Rebecca's code would not compile on my machine. So, in practice, her design contains compile-time errors (and possibly run-time errors). Nevertheless, her code provides ample evidence that she was attempting conditional logic to handle each possible day.} In practice, her code creates seven different conditional branches---one branch for each day of the week---where the code within each branch is duplicated.

@@

The two code blocks demonstrate just how much code is duplicated for handling user input based on days of the week. Between these two chunks there are only three differences (lines 1, 3, and 11): all of the references to day are changed from 5 (on the left) to 7 (on the right).\footnote{In the text-input files students were given, days of the week were represented as integers (rather than the perhaps more familiar ``Tuesday,'' ``Wednesday,'' etc.).} Moreover, the changes from block to block are patternistic and predictable: the first line of the block is a non-functioning comment, the third line of each block just checks whether the rest of the block should run, while the eleventh line of each block compares an array entry to the day of interest. Everything else is duplicate boilerplate that is essentially repeated 7 times; once for each day of the week. I say ``essentially repeated'' because, as we'll now explore, there are minute differences between some of the code chunk's seven incarnations.

@@

In the previous section, I described two unusual features of Rebecca's code for searching one-stop flights:

@@

In this section, I offer explanations of Rebecca's design choices by interpreting data from over five hours of clinical interviews I conducted with her. I draw from those interviews to explain how design decisions that might seem unusual to an expert in fact grew rather unproblematically (for Rebecca) as ways of deliberately transferring prior knowledge and designs (which explains feature 1) or coping with a constraint to produce a reliable solution she could trust (which explains feature 2).

@@

As Rebecca explained, something as simple as finding a flight from, say, JFK to BWI ``involves scanning through multiple files, because it's not like one file that has everything conveniently like, there'' (Interview, March 16, 2012). When I asked what would make things easier if, hypothetically, all the information she needed were in one file, Rebecca responded by appealing to a previous assignment from last semester. In Rebecca's first-semester programming course (Basic Programming for Engineers), one of several multi-week projects had students create a system for users to conduct a fantasy football draft. The project involved, among other things, topics related to basic file management in the C language, including how to read in a file from disk (Interview, March 16, 2012). Rebecca explained:

@@

Our interview with Rebecca corroborates the snapshot evidence that
Rebecca directly copied code from a prior project. Moreover, Rebecca's
behavior of copying her old fantasy football code suggests several
implications.

>>>>

What's common to these threads of research is how they mine the data. In
most applications of code snapshot\footnote{What we call ``code
  snapshot'' research may be more formally called ``online protocol''
  research \cite{jadud_methods_2006,rodrigo_analyzing_2009}. ``Online'' here isn't mean to mean
  the sense of globally-connected computers, but rather that student
  materials are collected in a minimally-intrusive way \emph{while
  students work}.} research, the aim is to average across events and
sessions to arrive at a characteristic measure of a student's behavior.
Jadud \cite{jadud_methods_2006}, for example, develops the idea of a student's ``Error
Quotient,'' a measure of how effectively a student addresses
compile-time errors in their code.\footnote{A useful way of thinking
  about Jadud's \cite{jadud_methods_2006} error quotient is that the computation believes
  in second chances. It doesn't penalize students for making errors; it
  penalizes students if they don't fix a given error before compiling
  the same piece of code \emph{again.}} Rodrigo and colleagues extend
the idea of error quotient and introduce similar
measures such as students' ``compilation profiles'' \cite{rodrigo_analyzing_2009}, ``frustration profiles'' \cite{rodrigo_affective_2009}, and ``error
profiles'' \cite{tabanao_predicting_2011}. Because these measures can be
collected in real-time as students code, they offer the potential to
identify at-risk students and respond with early
interventions.

@@

The data presented in this case are taken from an ongoing IRB-approved
study of undergraduate electrical engineering majors undertaken at
\emph{Flagship State}, a large, public research institution on the east
coast. For two semesters, we followed a total of 10 students taking
``Intermediate Programming Concepts for Engineers.'' It is the second of
a required two-semester course sequence in programming.\footnote{Hereafter,
  we use ``Intermediate Programming'' to refer to the second course and
  ``Basic Programming'' to refer to the first.} Students can (and some
do) place out of Basic Programming via AP Computer Science credit, but
all Electrical and Computer Engineering students must take Intermediate
Programming.

@@

Our analysis details the work of Rebecca, a female first-year electrical
engineering major. We focus specifically on Rebecca's code for finding
``one-stop'' flights, which the instructor defined as ``all pairs of
flights that route the user between the departure and arrival airports
with exactly 1 stop (i.e., a one-connection flight)'' (Flights Database
handout, 2012). The one-stop problem is particularly challenging. To
solve it successfully students' code must accept a user's choice of
airports and day, then stitch together routes that involve two separate
flights in a way that passes stringent constraints for acceptable
layover times.

@@

This study proceeds from an empirical challenge: how can we develop
accounts of students' programming activity that explain the form and
evolution of their code on a design project? The focused form of that
challenge for this study is ``how can we understand the unconventional
design choices embedded in Rebecca's one-stop flight code?'' To answer
that question, this study draws from three data streams: ethnographic
observation, clinical interviewing, and code snapshot analysis.

For two semesters, one author was ethnographically embedded in the same
instructor's section of Intermediate Programming. His aim throughout was
to see what students see in terms of course material, assignment
directives, and instruction. In fall 2011 he observed approximately 50\%
of the course lectures. He also independently completed all class
homeworks and three out of the four course projects to more fully
understand the course's assessments. In spring 2012 he continued
attending lectures, though less frequently, and began attending select
TA-led discussion sections. During both lectures and discussion sections
he took field notes while recording ambient audio using a LiveScribe
Pulse pen.

Rebecca was one of four students (three female, one male) willing and
able to participate in a series of 1-hour outside-of-class clinical
interviews during the spring 2012 semester.\footnote{In total, roughly
  30 students responded to the initial in-class solicitation to be
  contacted by email about the study. Of the students emailed,
  approximately 8 students responded to the emails to schedule interview
  times. Of those 8, only four students were able to successfully find
  interview time slots that fit our respective schedules.} Rebecca was interviewed
 five times, and typical interviews split time between
asking about her experiences in the course and giving her time in the
interview to work on her project code. During each interview,
simultaneously used were:

@@

The final component of data gathering is modeled after Jadud's (2006)
system for capturing students' code. We developed
software, built around the open-source version control system called
Git, that effectively creates an entire copy of a student's code---what
we call ``snapshots''---every time students invoke the compiler on their
code. Our software then sends those snapshots to a secure,
researcher-accessible server in real-time as they're created.
Consequently, he could plan each interview with Rebecca around
up-to-the-minute knowledge of her work---in some cases work she had
completed just hours before the interview---and tailor interview
questions to emerging patterns in her code. In total, Rebecca's work
resulted in 958 compilation snapshots over the course of the semester.

@@

In this phase of analysis, we restrict data to only the final version of \texttt{one\_stop\_flight.c} Rebecca submitted as part of her project. This restriction is important because of the data that gets left out. We're forced to see Rebecca's design work the way her instructor did when he graded it: as a final product. The submitted code contains little---if any---evidence of design iteration. And, we have almost no access to streams of activity (inscriptional, gestural, verbal, and otherwise material) that would help us understand Rebecca's early stage design \citep{petre_software_2014}. Indeed, other than the code itself, the only artifacts that actually carry Rebecca's voice in this phase of analysis are the comments she places in her code. Put another way: restricting our scope to just code---and only the final submitted code at that---denies the analyst access to the channels of ``talk, embodied action, and inscription'' involved in design work \citep[ p.~179]{hall_disrupting_2002}.

@@

While Rebecca's design isn't optimized for speed, it uses drastically less memory than do other solutions (including the instructor's official solution). What Rebecca could have done (and what we'll discuss in a later section) is read the entire contents of all data files into RAM. We'll call that approach an \emph{in-memory} solution, because all of the flight information is loaded into memory. Instead, her solution stores only a handful of lines (4 at most) in memory at a time, leaving the rest \emph{on-disk}. An analogy helps clarify the difference.

@@

For each of the four projects during the semester, the instructor gave students what he called a ``design brief.'' Each design brief outlined the problem to be solved as well as any constraints imposed on students' solutions. For example, in the flights project, students' programs had to reject candidate multi-stop trip routes if the layover time would be too short (under 30 minutes) or too long (more than two hours). But, in addition to what we might call \emph{user-centered constraints} (viz., reject multi-stop trips that would have grueling layovers), the instructor also directed students on \emph{implementation details}: ways their program should work at a technical level that would be invisible to the user.

@@

A second key feature of Rebecca's code is the almost identical repetition of a single 23-line code chunk seven times (lines 50--240). Because users can run queries by choosing a day to fly (and some flights only run on certain days), students' code must be able to handle each of the seven possible days for when a user would want to fly. In principle, Rebecca's code achieves just that.\footnote{We say ``in principle'' because Rebecca's code would not compile on our machine. So, in practice, her design contains compile-time errors (and possibly run-time errors). Nevertheless, her code provides ample evidence that she was attempting conditional logic to handle each possible day.} In practice, her code creates seven different conditional branches---one branch for each day of the week---where the code within each branch is duplicated.

@@

The two code blocks demonstrate just how much code is duplicated for handling user input based on days of the week. Between these two chunks there are only three differences (lines 1, 3, and 11): all of the references to day are changed from 5 (on the left) to 7 (on the right).\footnote{In the text-input files students were given, days of the week were represented as integers (rather than the perhaps more familiar ``Tuesday,'' ``Wednesday,'' etc.).} Moreover, the changes from block to block are patternistic and predictable: the first line of the block is a non-functioning comment, the third line of each block just checks whether the rest of the block should run, while the eleventh line of each block compares an array entry to the day of interest. Everything else is duplicate boilerplate that is essentially repeated 7 times; once for each day of the week. We say ``essentially repeated'' because, as we'll now explore, there are minute differences between some of the code chunk's seven incarnations.

@@

In the previous section, we described two unusual features of Rebecca's code for searching one-stop flights:

@@

In this section, we offer explanations of Rebecca's design choices by interpreting data from over five hours of clinical interviews conducted with her. We draw from those interviews to explain how design decisions that might seem unusual to an expert in fact grew rather unproblematically (for Rebecca) as ways of deliberately transferring prior knowledge and designs (which explains feature 1) or coping with a constraint to produce a reliable solution she could trust (which explains feature 2). Much of this section reflects a first-hand account of one author interviewing Rebecca and, so, is related in the first person.

@@

As Rebecca explained, something as simple as finding a flight from, say, JFK to BWI ``involves scanning through multiple files, because it's not like one file that has everything conveniently like, there'' (Interview, March 16, 2012). When asked what would make things easier if, hypothetically, all the information she needed were in one file, Rebecca responded by appealing to a previous assignment from last semester. In Rebecca's first-semester programming course (Basic Programming for Engineers), one of several multi-week projects had students create a system for users to conduct a fantasy football draft. The project involved, among other things, topics related to basic file management in the C language, including how to read in a file from disk (Interview, March 16, 2012). Rebecca explained:

@@

The interviews with Rebecca corroborate the snapshot evidence that
Rebecca directly copied code from a prior project. Moreover, Rebecca's
behavior of copying her old fantasy football code suggests several
implications.
